---
title: "PD VD T1 and Analysis
author: "Katie Lawler"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: null
---

#Packages
```{r}

library(dplyr)
library(ggplot2)
library(broom)
library(haven)  
library(purrr)
library(stringr)
library(table1)
library(flextable)
library(DiagrammeRsvg)
library(car)
library(officer)
pacman::p_load(tidyverse, readr, table1, DiagrammeR, rsvg)


```


#Importing Datasets
```{r}
demographics_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Data & Analysis/Initial Data (Uncleaned)/DemographicsData.xpt')

bmi_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Data & Analysis/Initial Data (Uncleaned)/BMIData.xpt')

diabetes_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Data & Analysis/Initial Data (Uncleaned)/DiabetesData.xpt')

vitamind_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Data & Analysis/Initial Data (Uncleaned)/VitaminDData.xpt')

gly_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Data & Analysis/Initial Data (Uncleaned)/GlyData.xpt')

alc_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Data & Analysis/Initial Data (Uncleaned)/AlcData.xpt')

kidney_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Data & Analysis/Initial Data (Uncleaned)/KidneyData.xpt')

physact_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Data & Analysis/Initial Data (Uncleaned)/PhysicalActData.xpt')

#All individual datasets at this point, just imported
```

#Merging Datasets Based on SEQN Number
```{r}
# Put them all in a list
datasets <- list(
  demographics_data,
  bmi_data,
  diabetes_data,
  vitamind_data,
  gly_data,
  alc_data,
  kidney_data,
  physact_data
)

# Merge into single dataset
initial_dataset <- reduce(datasets, full_join, by = "SEQN")

#Check
nrow(initial_dataset)

```

#Cleaning Prediabetes Var
```{r}
initial_dataset <- initial_dataset %>%
  mutate(
    prediabetes = case_when(
      DIQ160 == 1 ~ 1,   # Yes
      DIQ160 == 2 ~ 0,   # No
      TRUE ~ NA_real_
    )
  )

# Filter to just prediabetes dataset
prediabetes_dataset <- initial_dataset %>%
  filter(prediabetes == 1)
```

#Cleaning and Categorizing Vitamin D Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    vitd_status = case_when(
      LBXVIDMS < 30                    ~ "Deficient",
      LBXVIDMS >= 30 & LBXVIDMS < 50  ~ "Insufficient",
      LBXVIDMS >= 50                   ~ "Sufficient",
      TRUE                             ~ NA_character_
    ) %>% 
      factor(levels = c("Deficient", "Insufficient", "Sufficient"))
  )


```


#Cleaning Sex/Gender Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    sex = case_when(
      RIAGENDR == 1 ~ "Male",
      RIAGENDR == 2 ~ "Female",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("Male", "Female"))
  )
```


#Cleaning Age Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(age = RIDAGEYR)
```

#Cleaning Race/Ethnicity Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    # Step 1: Standard NHANES categories
    race_ethnicity = case_when(
      RIDRETH3 == 1 ~ "Mexican American",
      RIDRETH3 == 2 ~ "Other Hispanic",
      RIDRETH3 == 3 ~ "Non-Hispanic White",
      RIDRETH3 == 4 ~ "Non-Hispanic Black",
      RIDRETH3 == 6 ~ "Non-Hispanic Asian",
      RIDRETH3 == 7 ~ "Other/Multi-Racial",
      TRUE ~ NA_character_
    ),
    # Create Binary Race for stable interactions
    # Non-Hispanic White is the reference group (first in levels)
    race_binary = case_when(
      race_ethnicity == "Non-Hispanic White" ~ "Non-Hispanic White",
      is.na(race_ethnicity) ~ NA_character_,
      TRUE ~ "Non-White"
    ) %>% factor(levels = c("Non-Hispanic White", "Non-White"))
  )

# Verify the reference group (the first level shown is the reference)
levels(prediabetes_dataset$race_binary)
```

#Cleaning BMI Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(bmi = BMXBMI)
```


#Cleaning & Recategorizing Alc Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    alcohol_intake = case_when(
      ALQ170 %in% c(777, 999) ~ NA_character_,    # refused/don't know
      ALQ170 >= 15 | ALQ170 == 30 ~ "High",       # 15–20 times or "more than 20 times"
      ALQ170 < 15 ~ "Low",                        # 0–14 times
      TRUE ~ NA_character_
    ) %>% factor(levels = c("Low", "High"))
  )
```

#Cleaning Kidney Issues Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    kidney_issue = case_when(
      KIQ022 == 1 ~ "Yes",
      KIQ022 == 2 ~ "No",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("No", "Yes"))
  )
```

#Cleaning & Recategorizing Phys. Act. Var 
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    sedentary_status3 = case_when(
      PAD680 %in% c(7777, 9999) ~ NA_character_,
      PAD680 < 300               ~ "Low",
      PAD680 >= 300 & PAD680 < 600 ~ "Medium",
      PAD680 >= 600              ~ "High",
      TRUE                        ~ NA_character_
    ) %>% factor(levels = c("Low", "Medium", "High"))
  )
```


#Cleaning Gly Control/HbA1C Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    hba1c = LBXGH
  )
```

#Complete Count
```{r}
#Check col names
colnames(prediabetes_dataset)

#Make sure BMI and Age included
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    age = RIDAGEYR,
    bmi = BMXBMI
  )

# Variables required to be complete
vars_to_check <- c("prediabetes", "vitd_status", "hba1c", "sex", "age", "race_ethnicity", "bmi")

# Count rows with complete data for key vars
complete_rows <- prediabetes_dataset %>%
  filter(complete.cases(across(all_of(vars_to_check)))) %>%
  nrow()

# Total rows in dataset
total_rows <- nrow(prediabetes_dataset)

message("Total participants: ", total_rows)
message("Participants with complete data for key variables: ", complete_rows)
message("Proportion complete: ", round(complete_rows / total_rows, 3))
```

#New Clean Dataset  - REMOVED alc, Kidney, Phys. Act!!!
```{r}
clean_prediabetes_dataset <- prediabetes_dataset %>%
  dplyr::select(
    SEQN,            
    hba1c,
    vitd_status,
    age,
    sex,
    race_ethnicity,
    bmi
  )

# Drop rows with missing values
clean_prediabetes_dataset <- clean_prediabetes_dataset %>%
  filter(complete.cases(.))

# Check
nrow(clean_prediabetes_dataset)
```

#Figure 1
```{r}

fig1 <- grViz(diagram = "
digraph flowchart {

  graph [
    label = 'Figure 1. Flow diagram of participant selection',
    labelloc = t,
    fontsize = 18,
    fontname = Arial
  ]

  node [fontname = Arial, shape = rectangle, fontsize = 15]

  node1 [label = '@@1']
  node2 [label = '@@2']
  node3 [label = '@@3']
  node4 [label = '@@4']

  node1 -> node2 -> node3 -> node4
}
[1]: 'Starting number of records in NHANES merged dataset (n = 11,933)'
[2]: 'Excluding 11,015 individuals without a prediabetes diagnosis (n = 918)'
[3]: 'Excluding 230 individuals with missing data on key variables (n = 688)'
[4]: 'Final Analytic Sample (n = 688)'
")

svg <- export_svg(fig1)

rsvg_png(
  charToRaw(svg),
  file = "Figure1_Flowchart.png",
  width = 1600,
  height = 1200
)

```


#Table 1 / Descriptives
```{r}
library(table1)

# Ensure vitd_status is a factor with correct order
clean_prediabetes_dataset$vitd_status <- factor(clean_prediabetes_dataset$vitd_status,
                                                levels = c("Deficient", "Insufficient", "Sufficient"))

# Label vars
label(clean_prediabetes_dataset$hba1c) <- "HbA1c (%)"
label(clean_prediabetes_dataset$age) <- "Age (Years)"
label(clean_prediabetes_dataset$sex) <- "Sex"
label(clean_prediabetes_dataset$race_ethnicity) <- "Race/Ethnicity"
label(clean_prediabetes_dataset$bmi) <- "BMI (Kg/M²)"

# Table1 
table1(~ hba1c + age + sex + race_ethnicity + bmi | vitd_status,
       data = clean_prediabetes_dataset,
       caption = "Table 1. Descriptive Characteristics of Prediabetic Participants Stratified by Vitamin D Status",
       render.categorical = function(x) {
         render.default(x, render.missing = " (Missing)", footnote = FALSE)
       },
       render.continuous = function(x) {
         render.default(x, render.missing = " (Missing)")
       })

# For knitting to HTML
cat('<div style="font-family: Times New Roman;">')
table1(~ hba1c + age + sex + race_ethnicity + bmi | vitd_status,
       data = clean_prediabetes_dataset,
       caption = "Table 1. Descriptive Characteristics of Prediabetic Participants Stratified by Vitamin D Status")
cat('</div>')


```

#Exporting Table1 via Flextable - NOT PRINTING RN
```{r}
# Load required packages
library(table1)
library(flextable)
library(officer)
library(dplyr)

# Make sure vitd_status is a factor with correct order
clean_prediabetes_dataset$vitd_status <- factor(clean_prediabetes_dataset$vitd_status,
                                                levels = c("Deficient", "Insufficient", "Sufficient"))

# Label variables for nicer table output
label(clean_prediabetes_dataset$hba1c) <- "HbA1c (%)"
label(clean_prediabetes_dataset$age) <- "Age (Years)"
label(clean_prediabetes_dataset$sex) <- "Sex"
label(clean_prediabetes_dataset$race_ethnicity) <- "Race/Ethnicity"
label(clean_prediabetes_dataset$bmi) <- "BMI (Kg/M²)"

# Create the table1 object (summary)
t1 <- table1(~ hba1c + age + sex + race_ethnicity + bmi | vitd_status,
             data = clean_prediabetes_dataset,
             render.continuous = function(x) render.default(x, render.missing = " (Missing)"),
             render.categorical = function(x) render.default(x, render.missing = " (Missing)", footnote=FALSE))

# Convert table1 object to data frame
t1_df <- as.data.frame(t1)

# Create a flextable from the data frame
ft <- flextable(t1_df)

# Set font to Times New Roman
ft <- font(ft, fontname = "Times New Roman")

# Add a title to the table
ft <- add_header_lines(ft, values = "Table 1. Descriptive Characteristics of Prediabetic Participants Stratified by Vitamin D Status")

# Optional: Auto-adjust column widths for better formatting
ft <- autofit(ft)

# Export to Word
#doc <- read_docx()
#doc <- body_add_flextable(doc, ft)
#print(doc, target = "Table1_Descriptives.docx")

# The file Table1_Descriptives.docx will be in your working directory

```

#Filtering/Reordering Vars for Analysis
```{r}

# Filter dataset to complete cases only for these variables
regression_dataset <- prediabetes_dataset %>%
  filter(complete.cases(across(all_of(vars_to_check))))

# Set 'Sufficient' as the ref category for Vitamin D
regression_dataset$vitd_status <- relevel(as.factor(regression_dataset$vitd_status), ref = "Sufficient")

#Check
nrow(regression_dataset)

```


#Stat Tests for Baseline Char Differences for Contin. Vars
```{r}
#Checking Distribution - normal use ANOVA vs KS for not

#HB1AC
# Histogram by group
library(ggplot2)

ggplot(regression_dataset, aes(x = hba1c)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ vitd_status)

# Q-Q plots
ggplot(regression_dataset, aes(sample = hba1c)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ vitd_status)

#AGE
# Histogram
ggplot(regression_dataset, aes(x = age)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ vitd_status)

# Q-Q plot
ggplot(regression_dataset, aes(sample = age)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ vitd_status)

#BMI
# Histogram
ggplot(regression_dataset, aes(x = bmi)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ vitd_status)

# Q-Q plot
ggplot(regression_dataset, aes(sample = bmi)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~ vitd_status)


# HbA1C and BMI use ANOVA
# Age use KW Test

###TESTS


# HbA1c — ANOVA
aov_a1c <- aov(hba1c ~ vitd_status, data = regression_dataset)
summary(aov_a1c)

# Age — Kruskal-Wallis
kruskal.test(age ~ vitd_status, data = regression_dataset)

# BMI — ANOVA
aov_bmi <- aov(bmi ~ vitd_status, data = regression_dataset)
summary(aov_bmi)


```


#Stat Tests for Baseline Char Differences for Categ. Vars.
```{r}
cat_vars <- c("sex", "race_ethnicity")

for (var in cat_vars) {
  cat("\n=========================\n")
  cat("Variable:", var, "\n")
  
  tbl <- table(regression_dataset[[var]], regression_dataset$vitd_status)
  print(tbl)
  
  # Chi-square test
  chi <- chisq.test(tbl)
  print(chi)
  
  # If expected counts are small → Fisher's Exact
  if (any(chi$expected < 5)) {
    cat("\nUsing Fisher's Exact Test:\n")
    fisher <- fisher.test(tbl)
    print(fisher)
  }
}
```


#Checking Assumptions for MLR
```{r}
#1. Linearity -- PASSED
  # Residuals vs Fitted Plot -- look for a flat red line centered at zero
  plot(mvlinearreg_model, which = 1)

#2. Independence of Errors -- PASSED
  # Durbin-Watson Test - a value near 2 indicates independence; p > 0.05 is good
  library(car)
  durbinWatsonTest(mvlinearreg_model)

#3. Homoscedasticity (Equal Variance) -- PASSED
    # Scale-Location Plot (Visual) -- look for a flat red line with a random cloud of points
    plot(mvlinearreg_model, which = 3)
    
#4. Normality of Errors/Resid. -- PASSED, a few outliers (but checked their values)
    # Q-Q Plot -- points should fall strictly on the dashed line
    plot(mvlinearreg_model, which = 2)
    #Also checked the outliers from the plot, all of their A1C levels are clinically possible

#5. No Multicolinearity -- PASSED
  # Variance Inflation Factor (VIF) -- alues > 10 or 5 indicate high multicollinearity, low is good
  vif(mvlinearreg_model)
    

  
```


#START OF ANALYSIS!!!!!!!

#MLR Model
```{r}
# Model
mvlinearreg_model <- lm(hba1c ~ vitd_status + sex + age + race_ethnicity + bmi, 
                      data = regression_dataset)

# View the results
summary(mvlinearreg_model)
        
```

#Printing MLR Model Results / Nicer Table
```{r}
# Model results into a clean data frame
model_results <- tidy(mvlinearreg_model, conf.int = TRUE)

# Create the flextable
reg_table <- flextable(model_results) %>%
  # Round the numbers to 3 decimal places for neatness
  colformat_double(digits = 3) %>%
  # Rename headers to be professional
  set_header_labels(
    term = "Variable",
    estimate = "Coefficient (β)",
    std.error = "Std. Error",
    statistic = "t-value",
    p.value = "p-value",
    conf.low = "95% CI (Lower)",
    conf.high = "95% CI (Upper)"
  ) %>%
  # Bold the significant results (p < 0.05)
  bold(i = ~ p.value < 0.05, j = ~ p.value) %>%
  # Aesthetics
  font(fontname = "Times New Roman", part = "all") %>%
  autofit()

# View 
reg_table

# Export to Word 
doc <- read_docx()
doc <- body_add_flextable(doc, reg_table)
print(doc, target = "Regression_Results_Table.docx")
```




#Interaction Terms
```{r}
# Models how VD interacts with every other covariate 
interaction_model_vD <- lm(hba1c ~ vitd_status * (sex + age + race_ethnicity + bmi), 
                           data = regression_dataset)

#Print results
summary(interaction_model_vD)

```

#Printing Interaction Terms / Nicer Table
```{r}
# Interaction model to clean data frame
interaction_results <- tidy(interaction_model_vD, conf.int = TRUE)

# Clean up the variable names for the table
interaction_results$term <- gsub("vitd_status", "Vitamin D ", interaction_results$term)
interaction_results$term <- gsub("sex", "Sex: ", interaction_results$term)
interaction_results$term <- gsub("race_ethnicity", "Race: ", interaction_results$term)
interaction_results$term <- gsub("bmi", "BMI", interaction_results$term)
interaction_results$term <- gsub("age", "Age", interaction_results$term)
interaction_results$term <- gsub(":", " * ", interaction_results$term)

# Create the flextable
int_table <- flextable(interaction_results) %>%
  colformat_double(digits = 3) %>%
  set_header_labels(
    term = "Predictor (Main Effects & Interactions)",
    estimate = "Beta (β)",
    std.error = "SE",
    p.value = "p-value",
    conf.low = "Lower CI",
    conf.high = "Upper CI"
  ) %>%
  # Highlight the significant interaction you found (Sex * Vit D)
  bg(i = ~ p.value < 0.05, bg = "#f2f2f2") %>% 
  bold(i = ~ p.value < 0.05) %>%
  font(fontname = "Times New Roman", part = "all") %>%
  fontsize(size = 9, part = "all") %>% # Slightly smaller to fit interaction terms
  autofit()

# View and Export
int_table

doc_int <- read_docx()
doc_int <- body_add_flextable(doc_int, int_table)
print(doc_int, target = "Table3_Interaction_Results.docx")
```


