---
title: "CleanRRData"
author: "KatieLawler"
date: "`r Sys.Date()`"
output: html_document
---

#Packages
```{r}
# Install and load necessary packages
install.packages(c("dplyr", "ggplot2", "broom", "haven"))
library(dplyr)
library(ggplot2)
library(broom)
library(haven)  
library(purrr)
library(stringr)

```


#Importing Datasets
```{r}
demographics_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Research Report/Initial Data (Uncleaned)/DemographicsData.xpt')

bmi_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Research Report/Initial Data (Uncleaned)/BMIData.xpt')

diabetes_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Research Report/Initial Data (Uncleaned)/DiabetesData.xpt')

vitamind_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Research Report/Initial Data (Uncleaned)/VitaminDData.xpt')

gly_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Research Report/Initial Data (Uncleaned)/GlyData.xpt')

alc_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Research Report/Initial Data (Uncleaned)/AlcData.xpt')

kidney_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Research Report/Initial Data (Uncleaned)/KidneyData.xpt')

physact_data <- read_xpt('/Users/maryklawler/Desktop/Capstone Research Report/Initial Data (Uncleaned)/PhysicalActData.xpt')

#All individual datasets at this point, just imported
```

#Merging Datasets Based on SEQN Number
```{r}
# Put them all in a list
datasets <- list(
  demographics_data,
  bmi_data,
  diabetes_data,
  vitamind_data,
  gly_data,
  alc_data,
  kidney_data,
  physact_data
)

# Merge into single dataset
initial_dataset <- reduce(datasets, full_join, by = "SEQN")

```

#Cleaning Prediabetes Var
```{r}
initial_dataset <- initial_dataset %>%
  mutate(
    prediabetes = case_when(
      DIQ160 == 1 ~ 1,   # Yes
      DIQ160 == 2 ~ 0,   # No
      TRUE ~ NA_real_
    )
  )

# Step 2: Filter to just prediabetes dataset
prediabetes_dataset <- initial_dataset %>%
  filter(prediabetes == 1)
```

#Cleaning and Categorizing Vitamin D Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    vitd_status = case_when(
      LBXVIDMS < 30 ~ "Deficient",
      LBXVIDMS >= 30 & LBXVIDMS < 50 ~ "Insufficient",
      LBXVIDMS >= 50 ~ "Sufficient",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("Deficient", "Insufficient", "Sufficient"))
  )
```


#Cleaning Sex/Gender Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    sex = case_when(
      RIAGENDR == 1 ~ "Male",
      RIAGENDR == 2 ~ "Female",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("Male", "Female"))
  )
```


#Cleaning Age Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(age = RIDAGEYR)
```

#Cleaning Race/Ethnicity Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    race_ethnicity = case_when(
      RIDRETH3 == 1 ~ "Mexican American",
      RIDRETH3 == 2 ~ "Other Hispanic",
      RIDRETH3 == 3 ~ "Non-Hispanic White",
      RIDRETH3 == 4 ~ "Non-Hispanic Black",
      RIDRETH3 == 6 ~ "Non-Hispanic Asian",
      RIDRETH3 == 7 ~ "Other/Multi-Racial",
      TRUE ~ NA_character_
    ) %>% factor()
  )
```

#Cleaning BMI Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(bmi = BMXBMI)
```


#Cleaning & Recategorizing Alc Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    alcohol_intake = case_when(
      ALQ170 %in% c(777, 999) ~ NA_character_,    # refused/don't know
      ALQ170 >= 15 | ALQ170 == 30 ~ "High",       # 15–20 times or "more than 20 times"
      ALQ170 < 15 ~ "Low",                        # 0–14 times
      TRUE ~ NA_character_
    ) %>% factor(levels = c("Low", "High"))
  )
```

#Cleaning Kidney Issues Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    kidney_issue = case_when(
      KIQ022 == 1 ~ "Yes",
      KIQ022 == 2 ~ "No",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("No", "Yes"))
  )
```

#Cleaning & Recategorizing Phys. Act. Var 
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    sedentary_status3 = case_when(
      PAD680 %in% c(7777, 9999) ~ NA_character_,   # refused/don't know
      PAD680 < 300 ~ "Low",                        # less than 5 hours/day
      PAD680 >= 300 & PAD680 < 600 ~ "Medium",     # 5–10 hours/day
      PAD680 >= 600 ~ "High",                       # 10+ hours/day
      TRUE ~ NA_character_
    ) %>% factor(levels = c("Low", "Medium", "High"))
  )
```


#Cleaning Gly Control/HbA1C Var
```{r}
prediabetes_dataset <- prediabetes_dataset %>%
  mutate(
    hba1c = LBXGH
  )
```

#Complete Count
```{r}
# List of variables to check
vars_to_check <- c(
  "vitd_status",       # Predictor
  "hba1c",             # Outcome
  "sex",               # Confounders
  "age",
  "race_ethnicity",
  "bmi",
  "alcohol_intake",
  "kidney_issue",
  "sedentary_status3"
)

# Count rows with complete data
complete_rows <- prediabetes_dataset %>%
  filter(complete.cases(select(., all_of(vars_to_check)))) %>%
  nrow()

# Total rows in dataset
total_rows <- nrow(prediabetes_dataset)

message("Total participants: ", total_rows)
message("Participants with complete data for all variables: ", complete_rows)
message("Proportion complete: ", round(complete_rows / total_rows, 3))
```


#New Clean Dataset
```{r}
clean_prediabetes_dataset <- prediabetes_dataset %>%
  select(
    SEQN,            # Keep the ID
    hba1c,
    vitd_status,
    age,
    sex,
    race_ethnicity,
    bmi,
    alcohol_intake,
    kidney_issue,
    sedentary_status3
  )

# Optionally, drop rows with missing values for complete-case analysis
#clean_prediabetes_dataset <- clean_prediabetes_dataset %>%
  #filter(complete.cases(.))

```


#Table 1/Descriptives
#TO DO
